Cognate beginnings to bilingual lexical growth

### 2023-05-08 manuscript.qmd
Fix phonological transcriptions and render manuscript

---
title: "Lexical frequency mediates the facilitation effect of cognateness in bilingual lexical acquisition"

shorttitle: Frequency and cognateness in lexical acquisition

@@

# set ggplot theme and colour palette ------------------------------------------

theme_set(theme_custom())

bayesplot_theme_set(theme_custom()  +
                        theme(panel.grid = element_blank(),
                              axis.text = element_text(size = 5),
                              strip.text = element_text(size = 5),
                              plot.title = element_text(size = 8),
                              plot.subtitle = element_text(size = 8)))

clrs <- c("#003f5c", "#58508d", "#bc5090", "#ff6361", "#ffa600")

@@

color_scheme_set(rep(clrs[1], 6))

@@

# Introduction

@@

Most research on early word acquisition relies extensively on data from monolingual children, and is oblivious to the fact that a substantial proportion of the world population acquires more than one language from early ages [@grosjean2021extent]. Previous research on bilingual vocabulary acquisition pointed to bilingual toddlers knowing, on average, less words in each of their languages than their monolinguals peers, and also to both groups knowing a similar amount of words when the bilinguals' two languages are taken into account. @hoff2012dual found that English-Spanish bilingual toddlers in South Florida (United States) knew less words in English than monolinguals did. Both groups knew a similar total amount of words when both English and Spanish vocabularies were counted together, which substantiates the importance of collecting data on both languages when assessing bilinguals' communicative development. Other studies have provided converging evidence that bilinguals know a similar or even larger number of words than monolinguals when the two languages are aggregated [@oller2002language; @pearson1994patterns; @pearson1993lexical; @patterson2004comparing; @patterson2004bilingual; @smithson2014bilingualism; @petitto2001bilingual; @gonzalez-barrero2020bilingual]. These studies rely on bilingual samples whose languages are relatively distant in typology: English is a Germanic language, while Spanish is a Romance language, and both languages differ substantially at the grammatical or the phonological level. Little is known about vocabulary acquisition in bilinguals learning a pair of typologically closer languages (e.g., when both are Romance languages).

For a given set of concepts, bilingual children may be exposed to two distinct sets of word-forms: one in each language. These cross-language word pairs that share meaning are referred to as *translation equivalents*. Throughout the present study, we make the assumption that bilinguals form common semantic representations for both languages (e.g., the concept of a *chair* is shared for the English and Spanish words 'chair' and 'silla')^[The referential contexts in which bilingual children hear (and learn) both labels may not be identical. For instance, the English word *finger* refers to eight parts of the body (excluding thumbs and toes), whereas its Spanish translation *dedos* refers to twenty body parts (fingers, thumbs, and toes). These discrepancies are reflected, for instance, in adults' performance in lexical decision tasks [@chaouch-orozco2023are]. The fact that some translation pairs share more semantic features than others does not, however, inform whether bilinguals form common or separate semantic representations for translation equivalents. What English speakers understand for 'fingers' might not be identical to what Spanish speakers understand for 'dedos'. But when an English speaker learns Spanish, it seems implausible that they would form a separate semantic representation for what 'dedo' conveys at the conceptual level, compared to the possibility that they associate the new label to the semantic representation they already had formed for 'fingers'. Therefore, the lack of complete semantic alignment is not necessarily an argument against the existence of common semantic representations for translation equivalents. For simplicity, and in line with previous (formal and verbal) models of bilingual lexical processing [e.g., @shook2013bilingual; @dijkstra2019multilink; @dijkstra1998bia; @kroll2010revised], we assume common semantic representations for translation pairs.]. Depending on the similarity between both languages, the two sets of words may overlap in form in varying degrees. When both languages are typologically close, like Spanish and Catalan (both Romance languages), they are more likely to share a large amount of cognates (i.e., form-similar translation equivalents) than two typologically more distant languages, like Spanish and English (Romance and Germanic languages, respectively). For instance, in the presence of a door, a Spanish-Italian (or a Spanish-Catalan) bilingual might be exposed to the words *puerta* and *porta* (cognates), whereas a Spanish-English bilingual might hear *puerta* and *door* (non-cognates). It could be the case that mapping two phonologically similar labels (cognates like *puerta*-*porta*) onto the same referent is easier than doing the same with two phonologically dissimilar labels (non-cognates, like *puerta* and *door*). If cognates are easier to acquire than non-cognates, bilinguals learning two languages that share a high proportion of cognates should benefit more often from this facilitation effect than bilinguals learning two languages with a lower proportion of cognates, and should therefore show larger vocabulary sizes.

@floccia2018introduction provided evidence for such a facilitation effect of typological distance and vocabulary size. The authors collected vocabulary data on word comprehension and production from 372 24-month-old bilingual toddlers living in the United Kingdom who were learning English and an additional language. The additional language was one of 13 typologically diverse languages: Bengali, Cantonese Chinese, Dutch, French, German, Greek, Hindi/Urdu, Italian, Mandarin Chinese, Polish, Portuguese, Spanish and Welsh. The authors calculated the average phonological overlap between the words in each of these additional languages and their translation equivalents in English, which was taken as a proxy of the degree of *cognateness* between each pair of languages. Phonological overlap was measured by computing the Levenshtein distance between each cross-language pair of phonetic transcriptions. The Levenshtein distance is a metric that computes the distance between two strings of characters/phonemes by counting the smallest number of insertions, deletions and substitutions one of the strings has to undergo to become identical to the other [@levenshtein1966binary]. For instance, the Levenshtein distance for the translation equivalent *cat*-*kat* (/kæt/-/\textipa{kAt}/) in English and Dutch would be one: a single edition would be needed to make both phonological translation identical (replacing /æ/ by /\textipa{A}/). The same translation equivalent in English and Greek (*cat*-*$\gamma\acute{\alpha}\tau\alpha$*, /kæt/-/\textipa{Gata}/) would score three, as it requires at least three editions: substituting /k/ for /\textipa{G}/, substituting /æ/ for /a/ and adding another /a/. Levenshtein distance scores are divided by the length of the longest string to normalise the similarity scores between 0 and 1. Finally, the result is subtracted from one, to get the complementary, and therefore a measure of phonological similarity, instead of dissimilarity. Following the previous examples: $1/4 = 0.25$ for /kæt/-/\textipa{kAt}/, and $3/4 = 0.75$ for /kæt/-/\textipa{Gata}/. Finally, this variable was included as a predictor of participants' vocabulary sizes. Among other findings, Floccia and co-workers reported an increase in vocabulary size in the additional language (i.e., not English) associated with an increase in the average phonological similarity between the translation equivalents of each language pair. For example, English-Dutch bilinguals (0.2214 mean phonological overlap), were able to produce more Dutch words than English-Mandarin bilinguals (0.0197 mean phonological overlap) were able to produce in Mandarin. More recently, @blom2020crosslanguage and @bosma2019longitudinal reported similar results, providing converging evidence of for facilitatory effect of language distance on vocabulary size estimates.

Floccia et al. pointed to *parallel activation* as the underpinning mechanism behind their results. The parallel activation hypothesis stems from the language non-selective account of lexical access, which suggests that bilinguals activate both languages simultaneously during speech production or comprehension, even in fully monolingual environments. Evidence supporting the language-non selective account of lexical access is not restricted to production or the auditory modality, but rather extend to comprehension and translation, and to the visual modality (reading and signing) [@spivey1999cross; @hoshino2008cognate; @schwartz2007reading; @schwartz2006bilingual; @morford2011when; @vanhell2008sentence; @gimeno-martinez2021crosslinguistic; see @kroll2013multilingual for review]. One of the clearest pieces of evidence of parallel activation was provided by @costa2000cognate. In this study, Spanish monolinguals and Catalan-Spanish bilingual adults were asked to name pictures of common objects in Spanish. In half of the trials, the object labels were cognates in Spanish and Catalan (*árbol*-*arbre*, translations of *tree*), whereas in the other half of the trial labels were non-cognates (*mesa*-*taula*, translations of *table*), importantly such distinction was only relevant for bilinguals. Bilinguals named cognate pictures faster than non-cognate pictures, even after adjusting for the lexical frequency of the items. In contrast, Spanish monolinguals, who were unfamiliar with the Catalan translations of the Spanish words they uttered, showed equivalent naming times for the two types of stimuli. These results showed that bilinguals' Catalan phonology was activated during the production of Spanish words, facilitating the naming of cognate pictures.

Bilingual toddlers and children---still in the process to developing a mature lexicon---also show evidence of parallel activation during comprehension [@vonholzen2012language; @vonholzen2019impact; @bosma2020cognate; @poarch2012crosslanguage; @schroter2016orthographic; @poulin-dubois2013lexical]. Whether---and if so, how---the language non-selectivity shapes the early bilingual lexicon is still unclear. Previous work has pointed to an interactive account of lexical acquisition, in which toddlers' lexical representations in one language influence the acquisition of words in the other language. This cross-language facilitation is, in principle, possible since early ages: the bilingual infant lexicon consists in a substantial proportion of  translation equivalents: the presence of a word in the lexicon is a good predictor of its translation equivalent also being present [@bilson2015semantic; @tsui2022are; @pearson1995crosslanguage; @pearson1995crosslanguage]. In fact, translation equivalents may even be overrepresented in the early bilingual lexicon. @bilson2015semantic found that bilingual toddlers produced a higher proportion of translation equivalents than it would be expected if they learned two independent lexica simultaneously, which suggests that bilingual children preferentially acquired words for which they already knew a label in the other language. In the same line, @bosch2014first reported a similar overrepresentation of translation equivalents in Catalan-Spanish bilinguals: a high proportion of acquired Catalan words held an also acquired translation in Spanish. Finally, a recent investigation by @tsui2022area presented a quantitative model of bilingual vocabulary acquisition which showed a consistent preferential acquisition of translation equivalents by English-French bilinguals, with such preference being stronger at early stages of vocabulary growth. Overall, these findings suggest that bilingual children benefit from a semantic facilitation effect driven by the acquisition of translation equivalents.

It is possible that cognateness increases the amount of cross-language activation, facilitating early word acquisition. This could ultimately lead to children who are learning languages with a larger proportion of cognates (e.g., English and Dutch) showing larger vocabulary sizes, as found by Floccia et al. There is evidence of for phonological and semantic similarity facilitating word acquisition in monolinguals: infants are more likely to acquired new words that are semantically or phonologically more connected to other words, than words with poorer connectivity in either of those dimensions, such as words belonging to sparse phonological neighbourhoods [@hills2009longitudinal; @fourtassi2020growth; @laing2022phonological; @jones2019children; @storkel2004children]. If phonological similarity plays a cross-language facilitation role during bilingual word acquisition, cognate translation equivalents---which share both semantic and phonological similarity---should be acquired, on average, earlier that non-cognate translation equivalents---which share semantic similarity (e.g., are taxonomically related) but not phonological similarity.

@@

In the present investigation we approach the cognate facilitation on word learning by extending the conceptual framework of accumulator models of language acquisition to the bilingual case. Accumulator models formalise the widespread assumption that a child's vocabulary growth rate is a function of mainly---but not exclusively---the number of learning instances that a child has accrued with a particular word [see @kachergis2022standard for review]. A learning instance of a word consists, for example, in hearing such word-form (*flower*) in the presence of its referent (a flower). As infants receive linguistic input, they also accumulate opportunities to learn words until a threshold is crossed, an the word can be considered to be acquired. In line with this account, the frequency with which a child encounters a particular word-form in their speech input is a critical determiner of the age at which such word will be acquired. The more frequently [@hidaka2013computational; @mollica2017how]. For instance, words with high lexical frequency like 'flower' [`r format(floor(childes[childes$token=="flower",]$freq_million), big.mark = ",")` times per million words in the CHILDES corpus, @macwhinney2000childes] are more likely to be acquired earlier than words with low lexical frequency such as 'taxi' (`r format(floor(childes[childes$token=="taxi",]$freq_million), big.mark = ",")` times per million words) [e.g., @goodman2008does]: infants might accumulate learning experiences with 'flower' faster than with 'taxi'.

Previous accounts on vocabulary growth based on accumulator models are a convenient framework to study bilingual vocabulary growth, as bilinguals share fundamentally similar learning mechanisms with monolinguals. Developmental divergences between both groups can be understood as the result of the the negotiation between such mechanisms and the particular demands of a dual language input [e.g., @werker2005primir; @curtin2011bilingual]. One of such divergences is the fact that bilinguals' frequency of exposure to a word-form is mediated by their degree of exposure to their languages. Assuming that bilinguals and monolinguals are exposed to an equivalent amount of linguistic input, a Spanish-English bilingual who is exposed to Spanish 80% of the time and to English 20% of the time would need more time to accumulate the same experience with 'flower' and 'taxi' than an English monolingual would (who listens 100% of the time to English).

We suggest that the cognateness effect proposed by Floccia et al. and Mitchell et al. facilitates word acquisition by reducing the number of learning instances needed by a word to be acquired, and that this effect takes place *only* once the (cognate) translation of such word has been acquired. We hypothesise that both forms in the translation equivalent (e.g., *gat* and *gato* [Catalan and Spanish for 'cat'], for a Catalan-Spanish bilingual) follow their acquisition trajectory as they would for a monolingual child: accumulating learning instances, which occur a a rate proportional to the lexical frequency of the word-form and the amount of exposure to the language they belong to. When one of the word-forms is sufficiently acquired, the number of learning instances left needed for the acquisition of the other word-form is reduced.

This facilitation would occur as the result of the semantic and phonological overlap between both words. When the child has sufficient lexical knowledge of *gat*, and therefore there is some for-referent association in place, future presentations of the similar sounding translation *gato* in the presence of the same reference should lead to an increased likelihood of acquisition. In the case of a non-cognate translation equivalent, such facilitation would be unlikely. For instance, even if the child has already acquired the word *gos*, hearing its dissimilar Spanish translation *perro* would hardly lead to an increased probability of acquisition. In line with this account, only the later-acquired form of the translation equivalent would benefit from its cognate status. Given that children are more likely to acquire words from languages to which they are more often exposed to [@david2008individual; @cattani2014how; @thordardottir2006bilingual; @hoff2012dual], the acquisition of words in the language of lower exposure should, on average, be more susceptible to the effect of cognateness. For instance, if the previous child is exposed 80% of the time to English, and 20% of the time to Spanish, words in Spanish (especially low-frequency ones) would show a cognateness facilitation effect, with an earlier age of acquisition for cognates than for non-cognates.

{apafg-hypotheses} illustrates graphically a theoretical account of the acquisition of a Catalan-Spanish cognate translation equivalent, *gat*-*gato* [cat] whose word-forms share three phonemes, and of a non-cognate translation equivalent, *gos*-*perro* [dog] whose word-forms only share one phoneme. We assume the case of a Catalan-Spanish bilingual child who is exposed 65% of the time to Catalan, and 35% of the time to Spanish. On the top panels, we illustrate the cumulative sum of learning instances that a child accumulates across ages for the Catalan word-form (/gat/ or /gos/) and its Spanish translation (/perro/ or /gato/). We assume an arbitrary threshold of learning instances at which the child is considered to have acquired a word (solid grey line). Points and triangles show the age-of-acquisition at which the amount of learning instances reaches the threshold. We show the predicted age-of-acquisition under two hypothesis: (1) a cognateness facilitation effect that lowers the acquisition threshold after one of the word-forms (in this case Catalan /gat/) has been acquired, and (2) no cognateness facilitation, for which the acquisition of the second word-form occurs unconditionally of the acquisition of the first. On the bottom, we show the predicted probability of acquisition for the Catalan and Spanish words under each hypothesis, as it would be expected to observe in a vocabulary questionnaire like the one in the present study. We assume that such probability starts at 0% at birth, reaches 50% (and grows the fastest) at the age of acquisition, and reaches 100% at some points in later ages.

```{r apafg-hypotheses}
#| label: apafg-hypotheses
#| apa-cap: "Schematic representation of the hypothesised cognate facilitation effect on word acquisition."
#| out-width: 100%
knitr::include_graphics(normalizePath("assets/img/diagram.png"))
```

In line with this hypothesis, the size of the effect of linguistic similarity on vocabulary size that Floccia et al. reported was larger in the additional language (language of lower exposure) vocabulary than in English vocabulary (language of higher exposure). Most participants in their sample were English-dominant, meaning that their relative amount of exposure to English was larger than in the additional language. Therefore, participants may have on average learned the English word-form of translation equivalents earlier than the word-form in the additional language. Because of children's imbalance in exposure to words in English and in the additional language, the acquisition of English words by English-dominant participants might then benefit less frequently from their cognate status (the other word-form is unlikely to be available yet), while the acquisition of words in the additional language would benefit more frequently from their phonological similarity with the (more likely available) English translation.

@@

```{r}
#| label: items
#| echo: false
#| message: false
#| warning: false
#| include: false
n_categories <- nrow(distinct(bvq_data$pool, semantic_category))

@@

n_items_multi <- n_items_lemma |> 
    filter(!is_multiword) |> 
    filter(include, !is.na(wordbank_lemma)) |> 
    drop_na(sampa) |> 
    distinct(te, language)

>>>>

---
title: "Cognate beginnings to bilingual lexical growth"

shorttitle: "Cognate beginnings to bilingual lexical growth"

@@

# set ggplot theme and colour palette ------------------------------------------
theme_set(
    theme_minimal() +
        theme(
            panel.grid = element_line(colour = "grey",
                                      linetype = "dotted"),
            axis.line = element_line(colour = "black"),
            text = element_text(size = 12, colour = "black"),
            axis.text = element_text(colour = "black")
        )
)

bayesplot_theme_set(
        theme(
            axis.line = element_line(colour = "black"),
            text = element_text(size = 12, colour = "black"),
            panel.grid = element_blank(),
            axis.text = element_text(size = 5),
            strip.text = element_text(size = 5),
            plot.title = element_text(size = 8),
            plot.subtitle = element_text(size = 8)
        )
)

clrs <- c("#003f5c", "#58508d", "#bc5090", "#ff6361", "#ffa600")
color_scheme_set(rep(clrs[1], 6))

@@

@@

@@

Most research on early word acquisition relies extensively on data from monolingual children, and is oblivious to the fact that a substantial proportion of the world population acquires more than one language from early ages [@grosjean2021extent]. Previous research on bilingual vocabulary acquisition pointed to bilingual toddlers knowing, on average, less words in each of their languages than their monolinguals peers, and also to both groups knowing a similar amount of words when the bilinguals' two languages are taken into account. @hoff2012dual found that English-Spanish bilingual toddlers in South Florida (United States) knew less words in English than monolinguals did. Both groups knew a similar total amount of words when both English and Spanish vocabularies were counted together, which substantiates the importance of collecting data on both languages when assessing bilinguals' communicative development. Other studies have provided converging evidence that bilinguals know a similar or even larger number of words than monolinguals when the two languages are aggregated [@oller2002language; @pearson1994patterns; @pearson1993lexical; @patterson2004comparing; @patterson2004bilingual; @smithson2014bilingualism; @petitto2001bilingual; @gonzalez-barrero2020bilingual]. These studies rely on bilingual samples whose languages are relatively distant in typology: English is a Germanic language, while Spanish is a Romance language, and both languages differ substantially at the grammatical and the phonological level. Little is known about vocabulary acquisition in bilinguals learning a pair of typologically closer languages (e.g., when both are Romance languages).

One way typological distance may impact bilingual trajectories of language acquisition is the presence (or absence) of *cognates* in their linguistic input. Quite often, bilingual children encounter more than one label for the same referent: one in each language. For instance, an English-Spanish bilingual might hear both *dog* and *perro* in the presence of a dog We will refer to such cross-language synonyms as *translation equivalents*.^[Throughout the present study, we will make the assumption that bilinguals form a common semantic representations for each translation equivalent (e.g., the concept 'dog' is shared for the English and Spanish words *dog* and *perro*). The referential contexts in which bilingual children hear (and learn) both labels may not be identical. For instance, the English word *finger* refers to eight parts of the body (excluding thumbs and toes), whereas its Spanish translation *dedos* refers to twenty body parts (fingers, thumbs, and toes). These discrepancies are reflected, for instance, in adults' performance in lexical decision tasks [@chaouch-orozco2023are]. The fact that some translation pairs share more semantic features than others does not, however, inform whether bilinguals form common or separate semantic representations for translation equivalents. What English speakers understand for 'fingers' might not be identical to what Spanish speakers understand for 'dedos'. But when an English speaker learns Spanish, it seems implausible that they would form a separate semantic representation for what 'dedo' conveys at the conceptual level, compared to the possibility that they associate the new label to the semantic representation they already had formed for 'fingers'. Therefore, the lack of complete semantic alignment is not necessarily an argument against the existence of common semantic representations for translation equivalents. For simplicity, and in line with previous (formal and verbal) models of bilingual lexical processing [e.g., @shook2013bilingual; @dijkstra2019multilink; @dijkstra1998bia; @kroll2010revised], we assume common semantic representations for translation pairs.]. Sometimes, the word-forms involved in a translation equivalent may be phonologically similar. Such form-similar translation equivalents are referred to as cognates. While the previously mentioned translation pair *dog* and *perro* are non-cognates (both word-forms hardly share any phonemes), the translation pair *cat* and *gato* can be considered cognates. The amount of cognates shared by a pair of languages is dependent on their typological distance. When both languages are typologically close, like Spanish and Catalan (both Romance languages), they are more likely to share a large amount of cognates (i.e., form-similar translation equivalents) than two typologically more distant languages, like English and Spanish (Romance and Germanic languages, respectively). Therefore, bilingual children learning two typologically close languages may be exposed to a larger amount of cognates than those learning two typologically more distant languages. For instance, in the presence of a door, a Catalan-Spanish bilingual might be exposed to the words *porta* and *puerta* (cognate translation equivalent with the common meaning 'door'), whereas an English-Spanish bilingual might hear *door* and *puerta* (non-cognate translation equivalent).

It could be the case that mapping two phonologically similar labels (cognates like *porta*-*puerta*) onto the same referent is easier than doing the same with two phonologically dissimilar labels (non-cognates, like *door* and *puerta*). If cognates are easier to acquire than non-cognates, bilinguals learning two languages that share a high proportion of cognates should benefit more often from this facilitation effect than bilinguals learning two languages with a lower proportion of cognates, and should therefore show larger vocabulary sizes. @floccia2018introduction provided evidence for such a facilitation effect of typological distance and vocabulary size. The authors collected vocabulary data on word comprehension and production from 372 24-month-old bilingual toddlers living in the United Kingdom who were learning English and an additional language. The additional language was one of 13 typologically diverse languages: Bengali, Cantonese Chinese, Dutch, French, German, Greek, Hindi/Urdu, Italian, Mandarin Chinese, Polish, Portuguese, Spanish and Welsh. The authors calculated the average phonological overlap between the words in each of these additional languages and their translation equivalents in English, which was taken as a proxy of the degree of *cognateness* between each pair of languages. Phonological overlap was measured by computing the Levenshtein distance between each cross-language pair of phonetic transcriptions. The Levenshtein distance is a metric that computes the distance between two strings of characters/phonemes by counting the smallest number of insertions, deletions and substitutions one of the strings has to undergo to become identical to the other [@levenshtein1966binary]. For instance, the Levenshtein distance for the translation equivalent *cat*-*kat* (/\textipa{kæt}/-/\textipa{kAt}/) in English and Dutch would be one: a single edition would be needed to make both phonological translation identical (replacing /\textipa{æ}/ by /\textipa{A}/). The same translation equivalent in English and Greek (*cat*-*$\gamma\acute{\alpha}\tau\alpha$*, /\textipa{kæt}/-/\textipa{"Ga.ta}/) would score three, as it requires at least three editions: substituting /\textipa{k}/ for /\textipa{G}/, substituting /\textipa{\ae}/ for /\textipa{a}/ and adding another /\textipa{a}/. Levenshtein distance scores are divided by the length of the longest string to normalise the similarity scores between 0 and 1. Finally, the result is subtracted from one, to get the complementary, and therefore a measure of phonological similarity, instead of dissimilarity. Following the previous examples: $1/4 = 0.25$ for /\textipa{kæt}/-/\textipa{kAt}/, and $3/4 = 0.75$ for /\textipa{kæt}/-/\textipa{"Ga.ta}/. Finally, this variable was included as a predictor of participants' vocabulary sizes. Among other findings, Floccia and co-workers reported an increase in vocabulary size in the additional language (i.e., not English) associated with an increase in the average phonological similarity between the translation equivalents of each language pair. For example, English-Dutch bilinguals (0.2214 mean phonological overlap), were able to produce more Dutch words than English-Mandarin bilinguals (0.0197 mean phonological overlap) were able to produce in Mandarin. More recently, @blom2020crosslanguage and @bosma2019longitudinal reported similar results, providing converging evidence of for facilitatory effect of language distance on vocabulary size estimates.

Floccia et al. pointed to *parallel activation* as the underpinning mechanism behind their results. The parallel activation hypothesis stems from the language non-selective account of lexical access, which suggests that bilinguals activate both languages simultaneously during speech production or comprehension, even in fully monolingual environments. Evidence supporting the language-non selective account of lexical access is not restricted to production or the auditory modality, but rather extends to comprehension and translation, and to the visual modality (reading and signing) [@spivey1999cross; @hoshino2008cognate; @schwartz2007reading; @schwartz2006bilingual; @morford2011when; @vanhell2008sentence; @gimeno-martinez2021crosslinguistic; see @kroll2013multilingual for review]. One of the clearest pieces of evidence of parallel activation was provided by @costa2000cognate. In this study, Spanish monolinguals and Catalan-Spanish bilingual adults were asked to name pictures of common objects in Spanish. In half of the trials, the object labels were cognates in Spanish and Catalan (*árbol*-*arbre*, translations of *tree*), whereas in the other half of the trial labels were non-cognates (*mesa*-*taula*, translations of *table*), importantly such distinction was only relevant for bilinguals. Bilinguals named cognate pictures faster than non-cognate pictures, even after adjusting for the lexical frequency of the items. In contrast, Spanish monolinguals, who were unfamiliar with the Catalan translations of the Spanish words they uttered, showed equivalent naming times for the two types of stimuli. These results showed that bilinguals' Catalan phonology was activated during the production of Spanish words, facilitating the naming of cognate pictures.

Bilingual toddlers and children---still in the process to developing a mature lexicon---also show evidence of parallel activation during comprehension [@vonholzen2012language; @vonholzen2019impact; @bosma2020cognate; @poarch2012crosslanguage; @schroter2016orthographic; @poulin-dubois2013lexical]. Whether---and if so, how---the language non-selectivity shapes the early bilingual lexicon is still unclear. Previous work has pointed to an interactive account of lexical acquisition, in which toddlers' lexical representations in one language influence the acquisition of words in the other language. This cross-language facilitation is, in principle, possible since early ages: the bilingual infant lexicon consists in a substantial proportion of  translation equivalents: the presence of a word in the lexicon is a good predictor of its translation equivalent also being present [@bilson2015semantic; @tsui2022are; @pearson1995crosslanguage; @pearson1995crosslanguage]. In fact, translation equivalents may even be overrepresented in the early bilingual lexicon. @bilson2015semantic found that bilingual toddlers produced a higher proportion of translation equivalents than it would be expected if they learned two independent lexica simultaneously, which suggests that bilingual children preferentially acquired words for which they already knew a label in the other language. In the same line, @bosch2014first reported a similar overrepresentation of translation equivalents in Catalan-Spanish bilinguals: a high proportion of acquired Catalan words held an also acquired translation in Spanish. Finally, a recent investigation by @tsui2022are presented a quantitative model of bilingual vocabulary acquisition which showed a consistent preferential acquisition of translation equivalents by English-French bilinguals, with such preference being stronger at early stages of vocabulary growth. Overall, these findings suggest that bilingual children benefit from a semantic facilitation effect driven by the acquisition of translation equivalents.

In this scenario, where translation equivalents are pervasive in the bilingual lexicon since early ages, it is possible that cognateness increases the amount of cross-language activation, boosting early word acquisition. This could ultimately lead to children who are learning languages with a larger proportion of cognates (e.g., English and Dutch) showing larger vocabulary sizes, as found by Floccia et al. There is evidence of phonological and semantic similarity facilitating word acquisition in monolinguals: infants are more likely to acquired new words that are semantically or phonologically more connected to other words, than words with poorer connectivity in either of those dimensions, such as words belonging to sparse phonological neighbourhoods [@hills2009longitudinal; @fourtassi2020growth; @laing2022phonological; @jones2019children; @storkel2004children]. If phonological similarity plays a cross-language facilitation role during bilingual word acquisition, cognate translation equivalents---which share both semantic and phonological similarity---should be acquired, on average, earlier that non-cognate translation equivalents---which share semantic similarity (e.g., are taxonomically related) but not phonological similarity.

@@

In the present investigation we approach the cognate facilitation on word learning by extending the conceptual framework of accumulator models of language acquisition to the bilingual case. Accumulator models formalise the widespread assumption that a child's vocabulary growth rate is a function of mainly---but not exclusively---the number of learning instances that a child has accrued with a particular word [see @kachergis2022standard for review]. A learning instance of a word consists, for example, in hearing such word-form (*flower*) in the presence of its referent (a flower). As infants receive linguistic input, they also accumulate opportunities to learn words until a threshold is crossed, an the word can be considered to be acquired. The frequency with which a child encounters a particular word-form in their speech input is therefore a critical determiner of the age at which such word will be acquired, in part because they accumulate learning experiences with high-frequency words faster than with lower-frequency words [@hidaka2013computational; @mollica2017how]. Following this account, words with high lexical frequency like 'flower' [`r format(floor(childes[childes$token=="flower",]$freq_million), big.mark = ",")` times per million words in the CHILDES corpus, @macwhinney2000childes] are predicted to be acquired earlier in age than words with low lexical frequency such as 'taxi' (`r format(floor(childes[childes$token=="taxi",]$freq_million), big.mark = ",")` times per million words). We argue that previous accounts on vocabulary growth based on accumulator models are a convenient framework to study bilingual vocabulary growth. Bilinguals share fundamentally similar learning mechanisms with their monolinguals peers. Developmental divergences between both groups can be understood as the result of the the negotiation between such mechanisms and the particular demands of a dual language input [e.g., @werker2005primir; @curtin2011bilingual]. One of such divergences is the fact that bilinguals' frequency of exposure to a word-form is mediated by their degree of exposure to their languages. Assuming that bilinguals and monolinguals are exposed to an equivalent amount of linguistic input, a Spanish-English bilingual who is exposed to Spanish 80% of the time and to English 20% of the time would need more time to accumulate the same experience with 'flower' and 'taxi' than an English monolingual (who listens 100% of the time to English). would need. In the present study, we incorporate this feature of the bilingual linguistic input to an account of the cognateness effect on word acquisition.

We suggest that the cognateness effect proposed by Floccia et al. and Mitchell et al. facilitates word acquisition by reducing the number of learning instances needed by a word to be acquired, and that this effect takes place *only* once the (cognate) translation of such word has been acquired. We hypothesise that both forms in the translation equivalent (e.g., *gat* and *gato* [Catalan and Spanish for 'cat'], for a Catalan-Spanish bilingual) follow their acquisition trajectory as they would for a monolingual child. As the infant receives speech input, both word-forms accumulate learning instances, which occur a a rate proportional to the lexical frequency of each word-form and the amount of exposure to the language the word-form belongs to. When one of the word-forms has accumulated a certain amount of learning instances, it is acquired. After such event takes places, we hypothesise that the number of learning instances left needed for the acquisition of the other word-form (yet to be acquired) is reduced, resulting in an earlier age of acquisition that it would be expected if the acquisition of both word-forms was independent. We further suggest that this facilitation is proportional to the amount of phonological similarity (i.e., cognateness) between both word-forms, in such way that the facilitation is larger within similar-sounding translation equivalents, and negligible for form-dissimilar translation equivalents.

Consider the example of the Catalan-Spanish cognate translation equivalent *gat*-*gato* [cat] (whose word-forms share three phonemes), and of the non-cognate translation equivalent *gos*-*perro* [dog] whose word-forms only share one phoneme. When the child has sufficient lexical knowledge of *gat*, and therefore there is some form-referent association in place, future presentations of the similar-sounding translation *gato* in the presence of the same reference should lead to an increased likelihood of acquisition. In the case of a non-cognate translation equivalent, such facilitation would be unlikely. For instance, even if the child has already acquired the word *gos*, hearing its dissimilar Spanish translation *perro* would hardly lead to an increased probability of acquisition. In line with this account, only the later-acquired form of the translation equivalent would benefit from its cognate status. Given that children are more likely to acquire words from languages to which they are more often exposed to [@david2008individual; @cattani2014how; @thordardottir2006bilingual; @hoff2012dual], the acquisition of words in the language of lower exposure should, on average, be more susceptible to the effect of cognateness. For instance, if the previous child is exposed 80% of the time to English, and 20% of the time to Spanish, words in Spanish (especially low-frequency ones) would show a cognateness facilitation effect, with an earlier age of acquisition for cognates than for non-cognates.

{apafg-hypotheses} provides a graphic, theoretical illustration of this account. We assume the case of a Catalan-Spanish bilingual child who is exposed 65% of the time to Catalan, and 35% of the time to Spanish. On the top panels, we illustrate the cumulative sum of learning instances that a child accumulates across ages for the Catalan word-form (/\textipa{gat}/ or /\textipa{gos}/) and its Spanish translation (/\textipa{"pe.ro}/ or /\textipa{"ga.to}/). We assume an arbitrary, theoretical threshold for the learning instances at which the child is considered to have acquired a word (solid grey line). Points and triangles show the age-of-acquisition at which the amount of learning instances reaches the threshold. We show the predicted age-of-acquisition under two hypothesis: (1) a cognateness facilitation effect that lowers the acquisition threshold after one of the word-forms (in this case Catalan /\textipa{gat}/) has been acquired, and (2) no cognateness facilitation, for which the acquisition of the second word-form occurs unconditionally of the acquisition of the first. On the bottom, we show the predicted probability of acquisition for the Catalan and Spanish words under each hypothesis, as it would be expected to observe in a vocabulary questionnaire like the one in the present study. We assume that such probability starts at 0% at birth, reaches 50% (and grows the fastest) at the age of acquisition, and reaches 100% at some points in later ages.

```{r apafg-hypotheses}
#| label: apafg-hypotheses
#| apa-cap: "Schematic representation of the hypothesised cognate facilitation effect on word acquisition."
#| out-width: 100%
#| fig-width: 6
#| fig-height: 5
knitr::include_graphics(normalizePath("assets/img/diagram.png"))
```

This hypothesis provides a convenient account for the results in Floccia et al. The size of the effect of linguistic similarity on vocabulary size that Floccia et al. reported was larger in the additional language (language of lower exposure) vocabulary than in English vocabulary (language of higher exposure). Most participants in their sample were English-dominant, meaning that their relative amount of exposure to English was larger than in the additional language. Therefore, participants may have on average learned the English word-form of translation equivalents earlier than the word-form in the additional language. Because of children's imbalance in exposure to words in English and in the additional language, the acquisition of English words by English-dominant participants might then benefit less frequently from their cognate status (the other word-form is unlikely to be available yet), while the acquisition of words in the additional language would benefit more frequently from their phonological similarity with the (more likely available) English translation.

@@

```{r methods-data}
#| label: items
#| echo: false
#| message: false
#| warning: false
#| include: false
n_categories <- nrow(distinct(bvq_data$pool, semantic_category))

@@

n_items_multi <- n_items_lemma |> 
    filter(!is_multiword) |> 
    filter(include, !is.na(wordbank_lemma)) |> 
    drop_na(xsampa) |> 
    distinct(te, language)
